Linear instability underlying synchronous waving of marine grass in flow

### 2014-09-21 Paper/grass.tex
changed the order of mode 1 mode 2 description and explained shear layer scaling

In our model the balance between viscous forces and the drag force in the shear layer 
together with the continuity of shear stress across the grass tip leads to a shear layer thickness $\delta=(\Rey\tilde{N_g})^{-1/3}H$.
% and velocity $u \sim \delta U_0$ in the shear layer. 
Where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, and $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity. 

@@

We solve this eigenvalue problem and a threshold in $\Rey$ of O(100), above which the flow is unstable (Re($\sigma$)$>$0), emerges from the 
solution of \eqref{Orr-somerfield}. The dependence of this critical $\Rey$, and the corresponding fastest growing wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown 
in figure \ref{Re_vs_delta}, and is found to compare well with experimental observations\cite{Ghisal02}.
The threshold Reynolds number increases with the grass density, indicating a competition between the 
destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
Previous linear stability calculations for terrestrial grass do not exhibit a threshold flow condition 
either due to the exclusion of the vegetation drag in their model\cite{Raupach96}, or because of ad-hoc
assumptions such as for the mean velocity profile\cite{Raupach96,Delangre06}.

@@

Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in
the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, is shown in the inset in figure \ref{Re_vs_delta}b. 
The observed frequencies are associated with the peaks in velocity spectra, frequency of $\monami$ and frequency of 
vortex passage \cite{Ghisal02}, and compare well with the predicted frequencies. 
%Regardless of the assumptions, the introduction of vegetation drag is observed to damp the linear modes\cite{Delangre06}.

@@

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 2 has a threshold $\Rey \sim ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in figure \ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of \eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but $\Rey/\Ndg \sim O(1)$,.
Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the grass is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
This estimate simplifies \eqref{Orr-somerfield} to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,   &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, &\text{ for $y>\hg$}.
\end{split}
\end{align}
Since $\Rey/\Ndg$ is the only remaining parameter, the mode shape would converge in the aforementioned limit, in agreement with our numerical results for Mode 2 shown in figure \ref{Asymptotic_mode}.

We interpret Mode 2 as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the grass tip plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).

\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of grass density. 
Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The mode shapes approach each other for these small values of $\delta/H$ indicating that the large grass density asymptote is reached. Mode 1 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. 
Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. 
The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the grass height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}

On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.

This limit can be understood through asymptotically estimating the sizes of the terms in \eqref{Orr-somerfield} using $D\sim \delta^{-1}$, $\sigma \sim \delta^{-1}$, and $U\sim \delta$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
The terms balance when $\Rey \sim {\Ndg}^{1/2}$ (or $\Rey \sim H/\delta$), leading to a simplification of \eqref{Orr-somerfield} as
\begin{align}
 \begin{split}
  \text{blah for $y<\hg$} \\
  \text{blah for $y>\hg$.}
 \end{split}
\end{align}
Since, $\Rey/\Ndg^{1/2}$ is the only remaining parameter in this model, explaining the numerically observed asymptote (see figure \ref{Re_vs_delta}). (this language should be improved)
This analysis concludes that the mode structure is self-similar over the length scale $\delta$; the verification of this expectation is shown in figure \ref{Asymptotic_mode}, supporting this argument.

From these asymptotic behaviors, Mode 1 may be interpreted as the instability of the flow in the boundary layer, whereas Mode 2 may be understood as the instability on the scale of the water column. 
Table \ref{tab:comparison} compares the two modes to each other, and to the Kelvin-Helmholtz instability. 
Mode 1 appears to be closely related to the Kelvin Helmholtz mechanism, whereas Mode 2 arises purely from the interaction between the unvegetated water column and the flow through the vegetation. 
Vegetation drag plays a dominant role in the mechanism for both the modes, which distinguishes our analysis from the traditional Kelvin-Helmholtz instability. 
The appearance of vegetation drag parameter in the dominant balance of \eqref{Orr-somerfield} and the threshold criteria demonstrates its role in setting the threshold.

Similar phenomenon of large amplitude coherent oscillation of terrestrial canopies in atmospheric flow known as \textit{honami}\cite{Inoue56,Raupach96}, have also been observed.
A crucial difference between the atmospheric and aquatic flow is that the atmospheric flows are essentially unbounded vertically\cite{Vivoni98,Nepf00}. 
Another major difference between the two is the considerable difference of stiffness of canopies; terrestrial canopies tend to be much more rigid, whereas aquatic canopies are buoyant\cite{Vivoni98,Ghisal02}. 
Despite these differences, in the framework of our model, the limit of $\hg/H \ll 1$ while $\delta/\hg$ = constant can be used to represents the hydrodynamic instability for the terrestrial case with stiff grass blades. We find that in this case, the transition from Mode 1 to Mode 2 happens at such a large grass density, so as to make Mode 2 irrelevant. We recover the Kelvin-Helmholtz-like characteristics observed in the terrestrial case. 
\newline
The deviation of our model predictions from the observed may be attributed to the various simplification we have made in our model. 
In a real canopy, the drag coefficients are known to vary from bottom to tip of the grass blades\cite{Vivoni98,Nepf00}. 
The turbulent viscosity is also known to vary between the flow of unvegetated region and flow through the canopy\cite{Ghisal02}. 
Although an improved analysis of flow with these variation in the model might lead to a better agreement between the observed and the predicted quantities, these variation are not central to the mechanism leading to the instability of flow.
\newline
In conclusion, we note that the threshold flow condition observed in the field and in lab experiments arises due to the presence of vegetation drag. 
We show the origin of this threshold using a consistent continuum hydrodynamic stabiity analysis for the shear flow that develops as a result of the drag. 
Such an analysis not only explains quantitative observations, but also provides an understanding of the hydrodynamic instability mechanism. 
The extent of the structure of Mode 2 throughout the depth of the water column implies an efficient transport of suspended tracers compared to that due to the localized structure of Mode 1 near the canopy top. 
Our analysis also informs flow structure formation in many other related scenarios, such as flow over coral reefs, permeable sediments, etc. and therefore is expected to have a wider impact.

\bibliography{Grass}{}
\bibliographystyle{plain}
\end{document}

>>>>

In our model the balance between viscous forces and the drag force in the shear layer $(\mu u_{sl}/\delta^2 \sim \rho C_N d N_g u_{sl}^2)$ and the continuity of shear stress across the grass tip $(u_{sl}/\delta \sim U_0/H)$ leads to a shear layer thickness $\delta=(\Rey\tilde{N_g})^{-1/3}H$. Where $\Ndg = \left(C_N d H N_g\right)$ is the vegetation frontal area per bed area, $\Rey=\rho U_0 H/\mu$ is the Reynolds number of the flow, $u_{sl}$ is the velocity in shear layer and $U_0 = {(dP/dx)~H^2}/{\mu}$ is scale for velocity. 

@@

We solve this eigenvalue problem and a threshold in $\Rey$ of O(100), above which the flow is unstable (Re($\sigma$)$>$0), emerges from the 
solution of \eqref{Orr-somerfield}. The dependence of this critical $\Rey$, and the corresponding fastest growing wavenumber $k$, on $\delta/H$ and $\hg/H$ is shown 
in figure \ref{Re_vs_delta}, and is found to compare well with experimental observations\cite{Ghisal02}.
The threshold Reynolds number increases with the grass density, indicating a competition between the 
destabilizing shear in the flow, and the stabilizing effect of damping due to vegetation drag.
Previous linear stability calculations for terrestrial grass do not exhibit a threshold flow condition 
either due to the exclusion of the vegetation drag in their model\cite{Raupach96}, or because of ad-hoc
assumptions for the mean velocity profile\cite{Raupach96,Delangre06}.

@@

Comparison of frequency (Im$(\sigma)$) of the fastest growing mode with that of experimental observations of frequencies in
the lab scale experiments, for cases where the vegetation was sufficiently dense to be modeled by a continuum drag field, is also shown in the inset in figure \ref{Re_vs_delta}b. 
The observed frequencies are associated with the peaks in velocity spectra, frequency of $\monami$ and frequency of 
vortex passage \cite{Ghisal02}, and compare well with the predicted frequencies. 
%Regardless of the assumptions, the introduction of vegetation drag is observed to damp the linear modes\cite{Delangre06}.

@@

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits an asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold. 
This limit can be understood through asymptotically estimating the sizes of the terms in \eqref{Orr-somerfield} using $D\sim \delta^{-1}$, $\sigma \sim \delta^{-1}$, and $U\sim \delta$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
The terms balance when $\Rey \sim {\Ndg}^{1/2}$ (or $\Rey \sim H/\delta$), leading to a simplification of \eqref{Orr-somerfield} as
\begin{align}
 \begin{split}
  \text{blah for $y<\hg$} \\
  \text{blah for $y>\hg$.}
 \end{split}
\end{align}
Since, $\Rey/\Ndg^{1/2}$ is the only remaining parameter in this model, explaining the numerically observed asymptote (see figure \ref{Re_vs_delta}). (this language should be improved)
This analysis concludes that the mode structure is self-similar over the length scale $\delta$; the verification of this expectation is shown in figure \ref{Asymptotic_mode}, supporting this argument.

On the other hand Mode 2 has a threshold $\Rey \sim ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in figure \ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of \eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but $\Rey/\Ndg \sim O(1)$,.
Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the grass is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
This estimate simplifies \eqref{Orr-somerfield} to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,   &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, &\text{ for $y>\hg$}.
\end{split}
\end{align}
\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of grass density. 
Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The mode shapes approach each other for these small values of $\delta/H$ indicating that the large grass density asymptote is reached. Mode 1 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. 
Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. 
The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the grass height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}
Since $\Rey/\Ndg$ is the only remaining parameter, the mode shape would converge in the aforementioned limit, in agreement with our numerical results for Mode 2 shown in figure \ref{Asymptotic_mode}. 
We interpret Mode 2 as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the grass tip plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).
%On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.

%This limit can be understood through asymptotically estimating the sizes of the terms in \eqref{Orr-somerfield} using $D\sim \delta^{-1}$, $\sigma \sim \delta^{-1}$, and $U\sim \delta$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
%The terms balance when $\Rey \sim {\Ndg}^{1/2}$ (or $\Rey \sim H/\delta$), leading to a simplification of \eqref{Orr-somerfield} as
%\begin{align}
% \begin{split}
%  \text{blah for $y<\hg$} \\
%  \text{blah for $y>\hg$.}
% \end{split}
%\end{align}
%Since, $\Rey/\Ndg^{1/2}$ is the only remaining parameter in this model, explaining the numerically observed asymptote (see figure \ref{Re_vs_delta}). (this language should be improved)
%This analysis concludes that the mode structure is self-similar over the length scale $\delta$; the verification of this expectation is shown in figure \ref{Asymptotic_mode}, supporting this argument.
From these asymptotic behaviors, Mode 1 may be interpreted as the instability of the flow in the boundary layer, whereas Mode 2 may be understood as the instability on the scale of the water column. 
Table \ref{tab:comparison} compares the two modes to each other, and to the Kelvin-Helmholtz instability. 
Mode 1 appears to be closely related to the Kelvin Helmholtz mechanism, whereas Mode 2 arises purely from the interaction between the unvegetated water column and the flow through the vegetation. 
Vegetation drag plays a dominant role in the mechanism for both the modes, which distinguishes our analysis from the traditional Kelvin-Helmholtz instability. 
The appearance of vegetation drag parameter in the dominant balance of \eqref{Orr-somerfield} and the threshold criteria demonstrates its role in setting the threshold.
Similar phenomenon of large amplitude coherent oscillation of terrestrial canopies in atmospheric flow known as \textit{honami}\cite{Inoue56,Raupach96}, have also been observed.
A crucial difference between the atmospheric and aquatic flow is that the atmospheric flows are essentially unbounded vertically\cite{Vivoni98,Nepf00}. 
Another major difference between the two is the considerable difference of stiffness of canopies; terrestrial canopies tend to be much more rigid, whereas aquatic canopies are buoyant\cite{Vivoni98,Ghisal02}. 
Despite these differences, in the framework of our model, the limit of $\hg/H \ll 1$ while $\delta/\hg$ = constant can be used to represents the hydrodynamic instability for the terrestrial case with stiff grass blades. We find that in this case, the transition from Mode 1 to Mode 2 happens at such a large grass density, so as to make Mode 2 irrelevant. We recover the Kelvin-Helmholtz-like characteristics observed in the terrestrial case. 
The deviation of our model predictions from the observed may be attributed to the various simplification we have made in our model. 
In a real canopy, the drag coefficients are known to vary from bottom to tip of the grass blades\cite{Vivoni98,Nepf00}. 
The turbulent viscosity is also known to vary between the flow of unvegetated region and flow through the canopy\cite{Ghisal02}. 
Although an improved analysis of flow with these variation in the model might lead to a better agreement between the observed and the predicted quantities, these variation are not central to the mechanism leading to the instability of flow.
\newline\newline
In conclusion, we note that the threshold flow condition observed in the field and in lab experiments arises due to the presence of vegetation drag. 
We show the origin of this threshold using a consistent continuum hydrodynamic stabiity analysis for the shear flow that develops as a result of the drag. 
Such an analysis not only explains quantitative observations, but also provides an understanding of the hydrodynamic instability mechanism. 
The extent of the structure of Mode 2 throughout the depth of the water column implies an efficient transport of suspended tracers compared to that due to the localized structure of Mode 1 near the canopy top. 
Our analysis also informs flow structure formation in many other related scenarios, such as flow over coral reefs, permeable sediments, etc. and therefore is expected to have a wider impact.
\bibliography{Grass}{}
\bibliographystyle{plain}
\end{document}
